---
title: "Demographic Analysis: Japan & United States"
author: "Report"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{caption}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.align  = "center",
  out.width  = "100%"
)
```

```{r libraries}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(scales)
library(tidycensus)
library(tigris)
library(sf)
library(patchwork)
library(knitr)
library(kableExtra)
library(viridis)
library(RColorBrewer)
library(stringr)
library(purrr) 

options(tigris_use_cache = TRUE)
options(scipen = 999)

# ── Census API key ─────────────────────────────────────────────────────────────
# Replace the string below with your key, or set it once in .Renviron:
#   census_api_key("YOUR_KEY", install = TRUE)
census_api_key("cae79ee00ab828454493cdbc63e0d615494c2a5e", overwrite = TRUE, install = FALSE)
```

\newpage

# Part 1 – Analysis of Japanese Demographic Data

## Data Loading & Cleaning

```{r japan-load}
# ── Subdivisions (population by year) ─────────────────────────────────────────
subdiv_raw <- read_excel("japan-subdivisions.xlsx")

# Standardise column names and drop the trailing arrow column
subdiv <- subdiv_raw %>%
  select(Name, Status, Native, pop1995, pop2000, pop2005, pop2010, pop2015, pop2020) %>%
  mutate(across(starts_with("pop"), as.numeric))

# ── Areas / density ────────────────────────────────────────────────────────────
areas_raw <- read_excel("japan-prefecture-areas.xlsx")

# Row 0 is an embedded sub-header – drop it and rename columns
areas <- areas_raw %>%
  slice(-1) %>%
  select(
    Rank         = 1,
    Prefecture   = 2,
    Japanese     = 3,
    Area_km2     = 4,
    Population   = 7,
    Density      = 8,
    Capital      = 12
  ) %>%
  mutate(
    Rank       = as.integer(Rank),
    Area_km2   = as.numeric(Area_km2),
    Population = as.numeric(gsub(",", "", as.character(Population))),
    Density    = as.numeric(Density)
  )

# Quick sanity check
cat("Subdivisions:", nrow(subdiv), "rows |",
    "Unique statuses:", paste(unique(subdiv$Status), collapse = ", "), "\n")
cat("Areas:", nrow(areas), "rows\n")
```

## 1.1 Prefecture-Level Population Change (1995–2020)

```{r prefecture-pct-change}
# Filter to prefectures only
prefectures <- subdiv %>%
  filter(Status == "Prefecture") %>%
  mutate(
    pct_2000 = (pop2000 - pop1995) / pop1995 * 100,
    pct_2005 = (pop2005 - pop1995) / pop1995 * 100,
    pct_2010 = (pop2010 - pop1995) / pop1995 * 100,
    pct_2015 = (pop2015 - pop1995) / pop1995 * 100,
    pct_2020 = (pop2020 - pop1995) / pop1995 * 100
  )

# Summary table
pref_table <- prefectures %>%
  select(Name, pop1995, pct_2000, pct_2005, pct_2010, pct_2015, pct_2020) %>%
  arrange(desc(pct_2020))

kable(pref_table,
      digits  = 1,
      col.names = c("Prefecture", "Pop 1995",
                    "% Δ 2000", "% Δ 2005", "% Δ 2010", "% Δ 2015", "% Δ 2020"),
      caption = "Percentage population change relative to 1995 for all 47 prefectures",
      booktabs = TRUE,
      format   = "latex") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"),
                font_size = 7)
```

## 1.2 Time-Series Line Plot – % Change Relative to 1995

```{r prefecture-lineplot, fig.height=8}
# Pivot to long format for plotting
pref_long <- prefectures %>%
  select(Name, pct_2000, pct_2005, pct_2010, pct_2015, pct_2020) %>%
  pivot_longer(
    cols      = starts_with("pct_"),
    names_to  = "Year",
    values_to = "pct_change"
  ) %>%
  mutate(Year = as.integer(str_extract(Year, "\\d{4}")))

# Label only endpoints
endpoint_labels <- pref_long %>%
  filter(Year == 2020) %>%
  arrange(pct_change)

# Colour-code growth vs decline
pref_long <- pref_long %>%
  left_join(
    prefectures %>% select(Name, pct_2020),
    by = "Name"
  ) %>%
  mutate(direction = ifelse(pct_2020 >= 0, "Growth", "Decline"))

ggplot(pref_long, aes(x = Year, y = pct_change,
                      group = Name, colour = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_line(alpha = 0.65, linewidth = 0.55) +
  geom_point(data = filter(pref_long, Year == 2020),
             size = 1.4, alpha = 0.8) +
  geom_text_repel(
    data          = endpoint_labels,
    aes(label     = Name),
    size          = 2.2,
    direction     = "y",
    nudge_x       = 1,
    segment.size  = 0.2,
    max.overlaps  = 15,
    colour        = "grey20"
  ) +
  scale_colour_manual(values = c("Growth" = "#2166ac", "Decline" = "#d73027")) +
  scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020),
                     limits = c(2000, 2023)) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title    = "Percentage Population Change Relative to 1995",
    subtitle = "All 47 Japanese Prefectures · 2000–2020",
    x        = NULL,
    y        = "% Change from 1995 Baseline",
    colour   = NULL,
    caption  = "Source: Japan Statistics Bureau"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```

## 1.3 1995 Population vs. 1995–2020 % Change

```{r prefecture-scatter}
pref_scatter <- prefectures %>%
  mutate(
    label_flag = abs(pct_2020) > 8 | pop1995 > 5e6,
    pop_m      = pop1995 / 1e6
  )

# Pearson correlation
r <- cor(pref_scatter$pop1995, pref_scatter$pct_2020, use = "complete.obs")

ggplot(pref_scatter, aes(x = pop_m, y = pct_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = pct_2020 >= 0), size = 3.5, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30",
              linewidth = 0.7, linetype = "solid") +
  geom_text_repel(
    data         = filter(pref_scatter, label_flag),
    aes(label    = Name),
    size         = 3,
    box.padding  = 0.35,
    max.overlaps = 15
  ) +
  scale_colour_manual(values = c("TRUE" = "#2166ac", "FALSE" = "#d73027"),
                      labels = c("TRUE" = "Growth", "FALSE" = "Decline")) +
  scale_x_continuous(labels = label_number(suffix = "M")) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  annotate("text", x = Inf, y = Inf,
           label   = paste0("r = ", round(r, 3)),
           hjust   = 1.1, vjust = 1.5,
           size    = 4, colour = "grey20") +
  labs(
    title    = "1995 Population vs. 1995–2020 Percentage Change",
    subtitle = "47 Japanese Prefectures",
    x        = "1995 Population (millions)",
    y        = "% Change 1995–2020",
    colour   = NULL,
    caption  = paste0("Pearson r = ", round(r, 3),
                      "; OLS regression line with 95% CI")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

**Interpretation:** The correlation coefficient (r = `r round(r, 3)`) suggests
`r if(abs(r) < 0.2) "little to no" else if(abs(r) < 0.5) "a weak" else "a moderate"`
relationship between a prefecture's initial population size and its subsequent growth
trajectory over the 1995–2020 period.

## 1.4 City Size and Population Change

```{r city-analysis, fig.height=6}
cities <- subdiv %>%
  filter(str_detect(Status, "City")) %>%
  mutate(
    pct_2020 = (pop2020 - pop1995) / pop1995 * 100,
    size_cat = case_when(
      pop1995 < 30000                    ~ "Small  (<30k)",
      pop1995 >= 30000  & pop1995 < 100000 ~ "Medium (30k–100k)",
      pop1995 >= 100000 & pop1995 < 300000 ~ "Large  (100k–300k)",
      pop1995 >= 300000 & pop1995 < 700000 ~ "Major  (300k–700k)",
      TRUE                               ~ "Metro  (700k+)"
    ),
    size_cat = factor(size_cat,
                      levels = c("Small  (<30k)", "Medium (30k–100k)",
                                 "Large  (100k–300k)", "Major  (300k–700k)",
                                 "Metro  (700k+)"))
  ) %>%
  filter(!is.na(pct_2020))

# Summary statistics by size category
city_summary <- cities %>%
  group_by(size_cat) %>%
  summarise(
    n         = n(),
    median_pct = median(pct_2020, na.rm = TRUE),
    mean_pct  = mean(pct_2020, na.rm = TRUE),
    pct_grew  = mean(pct_2020 > 0, na.rm = TRUE) * 100,
    .groups   = "drop"
  )

kable(city_summary,
      digits = 1,
      col.names = c("City Size", "N", "Median % Δ", "Mean % Δ", "% Cities that Grew"),
      caption  = "Population change 1995–2020 by city size category",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))

# Box plot
ggplot(cities, aes(x = size_cat, y = pct_2020, fill = size_cat)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(alpha = 0.8, outlier.size = 1.2, outlier.alpha = 0.5) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title    = "Population Change 1995–2020 by City Size",
    subtitle = "Japanese Cities (shi)",
    x        = "City Size Category (1995 Population)",
    y        = "% Change 1995–2020",
    fill     = NULL,
    caption  = paste0("n = ", nrow(cities), " cities")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1),
        panel.grid.minor = element_blank())
```

## 1.5 Population Density vs. % Population Growth (Prefectures)

```{r density-correlation, fig.height=6}
# Join prefectures with areas data
pref_density <- prefectures %>%
  left_join(areas %>% select(Prefecture, Density),
            by = c("Name" = "Prefecture")) %>%
  filter(!is.na(Density))

r_dens <- cor(pref_density$Density, pref_density$pct_2020, use = "complete.obs")

cat("Pearson correlation (Density vs % Growth):", round(r_dens, 4), "\n")
cat("Spearman correlation (Density vs % Growth):",
    round(cor(pref_density$Density, pref_density$pct_2020,
              method = "spearman", use = "complete.obs"), 4), "\n")

ggplot(pref_density, aes(x = Density, y = pct_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = pct_2020 >= 0), size = 3.5, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30", linewidth = 0.7) +
  geom_text_repel(
    aes(label = Name),
    size = 3, max.overlaps = 12,
    box.padding = 0.3
  ) +
  scale_colour_manual(values = c("TRUE" = "#2166ac", "FALSE" = "#d73027"),
                      labels = c("TRUE" = "Growth", "FALSE" = "Decline")) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  annotate("text", x = Inf, y = Inf,
           label  = paste0("Pearson r = ", round(r_dens, 3)),
           hjust  = 1.1, vjust = 1.5, size = 4, colour = "grey20") +
  labs(
    title    = "Population Density vs. Population Change 1995–2020",
    subtitle = "47 Japanese Prefectures (density on log scale)",
    x        = "Population Density 1995 (persons/km²) — log scale",
    y        = "% Change 1995–2020",
    colour   = NULL,
    caption  = paste0("Source: Japan Statistics Bureau / japan-prefecture-areas.xlsx\n",
                      "Pearson r = ", round(r_dens, 3),
                      "  |  Spearman ρ = ",
                      round(cor(pref_density$Density, pref_density$pct_2020,
                                method = "spearman", use = "complete.obs"), 3))
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

\newpage

# Part 2 – U.S. Census & TidyCensus Analysis

## 2.1 Variable Verification

Before pulling data, we document the exact Census variables used.

| Variable     | Decennial Year | Table | Description |
|:-------------|:------:|:------|:------------|
| `P001001`    | 2000   | P1    | Total population (SF1, Summary File 1) |
| `P1_001N`    | 2020   | P1    | Total population (Redistricting data / PL 94-171) |

```{r variable-check}
# Inspect 2000 and 2020 variable catalogues
vars_2000 <- load_variables(2000, "sf1", cache = TRUE)
vars_2020 <- load_variables(2020, "pl",  cache = TRUE)

vars_2000 %>% filter(name == "P001001") %>%
  select(name, label, concept) %>%
  kable(caption = "2000 Census variable P001001", booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")

vars_2020 %>% filter(name == "P1_001N") %>%
  select(name, label, concept) %>%
  kable(caption = "2020 Census variable P1\\_001N", booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

## 2.2 County Population Change 2000–2020

```{r county-pop, cache=TRUE}
# Pull 2000 county populations
pop_2000 <- get_decennial(
  geography = "county",
  variables = "P001001",
  year      = 2000,
  sumfile   = "sf1"
) %>%
  select(GEOID, NAME, pop2000 = value)

# Pull 2020 county populations
pop_2020 <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  year      = 2020,
  sumfile   = "pl"
) %>%
  select(GEOID, NAME, pop2020 = value)

# Combine and calculate % change
county_pop <- pop_2000 %>%
  inner_join(pop_2020, by = "GEOID") %>%
  mutate(
    pct_change = (pop2020 - pop2000) / pop2000 * 100,
    state_fips = str_sub(GEOID, 1, 2)
  ) %>%
  filter(pop2000 > 0)

cat("Counties with population data:", nrow(county_pop), "\n")
cat("Summary of % change:\n")
print(summary(county_pop$pct_change))
```

## 2.3 Loading RUCC Codes

```{r rucc-load}
# USDA Rural-Urban Continuum Codes 2013
# Download manually from:
#   https://www.ers.usda.gov/data-products/rural-urban-continuum-codes/
# Save as "ruralurbancodes2013.xls" in the working directory.
# Column layout: FIPS | State | County_Name | Population_2010 | RUCC_2013 | Description

rucc_path <- "ruralurbancodes2013.xls"

if (file.exists(rucc_path)) {
  rucc_raw <- read_excel(rucc_path)

  # Flexible column detection
  fips_col  <- grep("(?i)fips",     names(rucc_raw), value = TRUE)[1]
  rucc_col  <- grep("(?i)rucc_2013", names(rucc_raw), value = TRUE)[1]

  rucc <- rucc_raw %>%
    rename(GEOID = all_of(fips_col),
           RUCC  = all_of(rucc_col)) %>%
    mutate(
      GEOID = str_pad(as.character(GEOID), 5, pad = "0"),
      RUCC  = as.integer(RUCC),
      metro = ifelse(RUCC <= 3, "Metro", "Nonmetro"),
      metro3 = case_when(
        RUCC %in% 1:3 ~ "Metro",
        RUCC %in% 4:6 ~ "Micropolitan",
        RUCC %in% 7:9 ~ "Rural"
      )
    ) %>%
    select(GEOID, RUCC, metro, metro3)

  county_pop <- county_pop %>% left_join(rucc, by = "GEOID")
  cat("RUCC file loaded. Distribution:\n")
  print(table(county_pop$RUCC, useNA = "ifany"))
} else {
  warning("RUCC file not found. Scatter plots coloured by RUCC will be skipped.")
  county_pop <- county_pop %>%
    mutate(RUCC = NA_integer_, metro = NA_character_, metro3 = NA_character_)
}
```

## 2.4 Scatter Plots – County Growth Rate vs. Total Population

### 2.4a All Counties (no grouping)

```{r scatter-all, fig.height=6}
ggplot(county_pop, aes(x = pop2000 / 1e3, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.25, size = 0.9, colour = "#2166ac") +
  geom_smooth(method = "loess", se = FALSE, colour = "#d73027",
              linewidth = 0.8, span = 0.5) +
  scale_x_log10(labels = label_comma(suffix = "k")) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     limits = c(-80, 200)) +
  labs(
    title    = "U.S. County Population Growth Rate vs. Population Size",
    subtitle = "2000–2020 · n ≈ 3,100 counties",
    x        = "2000 Population (thousands) — log scale",
    y        = "% Change 2000–2020",
    caption  = "Source: U.S. Decennial Census via tidycensus"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

### 2.4b Coloured by RUCC Code

```{r scatter-rucc, fig.height=6}
if (!all(is.na(county_pop$RUCC))) {
  rucc_palette <- colorRampPalette(c("#1a9641", "#ffffbf", "#d7191c"))(9)

  ggplot(county_pop %>% filter(!is.na(RUCC)),
         aes(x = pop2000 / 1e3, y = pct_change,
             colour = factor(RUCC))) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(alpha = 0.35, size = 0.9) +
    scale_colour_manual(
      values = rucc_palette,
      name   = "RUCC 2013",
      labels = c("1 Metro ≥1M", "2 Metro 250k–1M", "3 Metro <250k",
                 "4 Adj+Urban 20k+", "5 Nonadj+Urban 20k+",
                 "6 Adj+Urban 2.5k–20k", "7 Nonadj+Urban 2.5k–20k",
                 "8 Adj+Rural", "9 Nonadj+Rural")
    ) +
    scale_x_log10(labels = label_comma(suffix = "k")) +
    scale_y_continuous(labels = label_number(suffix = "%"),
                       limits = c(-80, 200)) +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2),
                                 ncol = 1)) +
    labs(
      title    = "County Growth Rate vs. Population – Coloured by RUCC",
      subtitle = "U.S. counties 2000–2020",
      x        = "2000 Population (thousands) — log scale",
      y        = "% Change 2000–2020",
      caption  = "RUCC = USDA Rural-Urban Continuum Codes 2013"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position   = "right",
          legend.text       = element_text(size = 7),
          panel.grid.minor  = element_blank())
} else {
  cat("RUCC data unavailable – plot skipped.\n")
}
```

### 2.4c Metro vs. Nonmetro (RUCC 1–3 vs. 4–9)

```{r scatter-metro, fig.height=6}
if (!all(is.na(county_pop$metro))) {
  ggplot(county_pop %>% filter(!is.na(metro)),
         aes(x = pop2000 / 1e3, y = pct_change, colour = metro)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(alpha = 0.3, size = 0.9) +
    geom_smooth(aes(group = metro), method = "loess",
                se = FALSE, linewidth = 1, span = 0.6) +
    scale_colour_manual(values = c("Metro" = "#2166ac", "Nonmetro" = "#d73027")) +
    scale_x_log10(labels = label_comma(suffix = "k")) +
    scale_y_continuous(labels = label_number(suffix = "%"),
                       limits = c(-80, 200)) +
    labs(
      title    = "County Growth Rate vs. Population – Metro vs. Nonmetro",
      subtitle = "Metro = RUCC 1–3  |  Nonmetro = RUCC 4–9",
      x        = "2000 Population (thousands) — log scale",
      y        = "% Change 2000–2020",
      colour   = NULL,
      caption  = "Source: U.S. Decennial Census & USDA ERS RUCC 2013"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position  = "top",
          panel.grid.minor = element_blank())
} else {
  cat("Metro classification unavailable – plot skipped.\n")
}
```

## 2.5 Population Pyramids – Illinois Counties

### Helper: Age/Sex Variables

```{r pyramid-vars}
# ACS 5-year 2020 – Table B01001 (Sex by Age)
acs_vars <- load_variables(2020, "acs5", cache = TRUE)

male_vars <- acs_vars %>%
  filter(str_detect(name, "^B01001_0") & str_detect(label, "Male")) %>%
  filter(!str_detect(label, "Female")) %>%
  pull(name)

female_vars <- acs_vars %>%
  filter(str_detect(name, "^B01001_0") & str_detect(label, "Female")) %>%
  pull(name)

# Standard 18 male age groups: B01001_003 – B01001_025 (skip _002 = total male)
male_vars   <- paste0("B01001_", str_pad(3:25, 3, pad = "0"))
female_vars <- paste0("B01001_", str_pad(27:49, 3, pad = "0"))

age_labels <- c(
  "Under 5", "5-9", "10-14", "15-17", "18-19", "20",
  "21", "22-24", "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-61", "62-64", "65-66",
  "67-69", "70-74", "75-79", "80-84", "85+"
)

# Broader 9-group labels for readability
broad_labels <- c(
  "<5","5–9","10–14","15–17","18–19","20","21","22–24",
  "25–29","30–34","35–39","40–44","45–49","50–54",
  "55–59","60–61","62–64","65–66","67–69","70–74",
  "75–79","80–84","85+"
)
```

### Illinois: Metro vs. Nonmetro

```{r illinois-pyramid, fig.height=7, cache=TRUE}
# Pull ACS age/sex for all Illinois counties
il_vars <- c(male_vars, female_vars)

il_acs <- get_acs(
  geography = "county",
  state     = "IL",
  variables = il_vars,
  year      = 2020,
  survey    = "acs5"
)

# Attach RUCC classification for Illinois
il_rucc <- county_pop %>%
  filter(str_sub(GEOID, 1, 2) == "17") %>%   # Illinois FIPS = 17
  select(GEOID, metro)

il_pyramid_data <- il_acs %>%
  left_join(il_rucc, by = "GEOID") %>%
  filter(!is.na(metro)) %>%
  mutate(
    sex       = ifelse(variable %in% male_vars, "Male", "Female"),
    age_idx   = ifelse(sex == "Male",
                       match(variable, male_vars),
                       match(variable, female_vars)),
    age_group = broad_labels[age_idx]
  ) %>%
  group_by(metro, sex, age_group, age_idx) %>%
  summarise(pop = sum(estimate, na.rm = TRUE), .groups = "drop") %>%
  group_by(metro) %>%
  mutate(pct = pop / sum(pop) * 100) %>%
  ungroup() %>%
  mutate(
    pct      = ifelse(sex == "Male", -pct, pct),
    age_group = factor(age_group, levels = broad_labels)
  )

# Build side-by-side pyramids
metro_pyramid <- function(df, title_suffix) {
  ggplot(df, aes(x = pct, y = age_group, fill = sex)) +
    geom_col(width = 0.8, alpha = 0.85) +
    geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.4) +
    scale_fill_manual(values = c("Male" = "#4393c3", "Female" = "#d6604d")) +
    scale_x_continuous(
      labels = function(x) paste0(abs(round(x, 1)), "%"),
      limits = c(-6.5, 6.5)
    ) +
    labs(
      title = title_suffix,
      x = "% of total county group population",
      y = NULL, fill = NULL
    ) +
    theme_minimal(base_size = 9) +
    theme(
      legend.position  = "top",
      panel.grid.minor = element_blank()
    )
}

p_metro    <- metro_pyramid(filter(il_pyramid_data, metro == "Metro"),
                            "Metro Counties")
p_nonmetro <- metro_pyramid(filter(il_pyramid_data, metro == "Nonmetro"),
                            "Nonmetro Counties")

(p_metro | p_nonmetro) +
  plot_annotation(
    title    = "Population Pyramids – Illinois Counties",
    subtitle = "Metro (RUCC 1–3) vs. Nonmetro (RUCC 4–9) · ACS 2020 5-year estimates",
    caption  = "Source: U.S. Census Bureau ACS 2020 via tidycensus"
  )
```

### Midwest: Metro, Micropolitan & Rural (Three-way Comparison)

```{r midwest-pyramid, fig.height=8, cache=TRUE}
midwest_fips <- c("17","18","19","20","26","27","29","31","38","39","46","55")
# IL IN IA KS MI MN MO NE ND OH SD WI

mw_acs <- get_acs(
  geography = "county",
  state     = midwest_fips,
  variables = il_vars,
  year      = 2020,
  survey    = "acs5"
)

mw_rucc <- county_pop %>%
  filter(str_sub(GEOID, 1, 2) %in% midwest_fips) %>%
  select(GEOID, metro3)

mw_pyramid_data <- mw_acs %>%
  left_join(mw_rucc, by = "GEOID") %>%
  filter(!is.na(metro3)) %>%
  mutate(
    sex       = ifelse(variable %in% male_vars, "Male", "Female"),
    age_idx   = ifelse(sex == "Male",
                       match(variable, male_vars),
                       match(variable, female_vars)),
    age_group = broad_labels[age_idx]
  ) %>%
  group_by(metro3, sex, age_group, age_idx) %>%
  summarise(pop = sum(estimate, na.rm = TRUE), .groups = "drop") %>%
  group_by(metro3) %>%
  mutate(pct = pop / sum(pop) * 100) %>%
  ungroup() %>%
  mutate(
    pct       = ifelse(sex == "Male", -pct, pct),
    age_group = factor(age_group, levels = broad_labels),
    metro3    = factor(metro3, levels = c("Metro", "Micropolitan", "Rural"))
  )

p_mw_metro <- metro_pyramid(filter(mw_pyramid_data, metro3 == "Metro"),
                             "Metro")
p_mw_micro <- metro_pyramid(filter(mw_pyramid_data, metro3 == "Micropolitan"),
                             "Micropolitan")
p_mw_rural <- metro_pyramid(filter(mw_pyramid_data, metro3 == "Rural"),
                             "Rural")

(p_mw_metro | p_mw_micro | p_mw_rural) +
  plot_annotation(
    title    = "Population Pyramids – Midwest U.S.",
    subtitle = "Metro · Micropolitan · Rural  |  ACS 2020 5-year estimates",
    caption  = "Midwest = IL, IN, IA, KS, MI, MN, MO, NE, ND, OH, SD, WI\nSource: U.S. Census Bureau ACS 2020 via tidycensus"
  )
```

## 2.6 Choropleth Map – Midwest County Population Change 2000–2020

```{r choropleth-midwest, fig.height=8, cache=TRUE}
# Download county shapefiles for Midwest states
#mw_counties <- map_dfr(
 # midwest_fips,
  #~ counties(state = .x, cb = TRUE, year = 2020, class = "sf")
#)

all_counties <- counties(cb = TRUE, year = 2020, class = "sf")
mw_counties  <- all_counties %>% filter(STATEFP %in% midwest_fips)

# Join with population change data
mw_map <- mw_counties %>%
  left_join(county_pop %>% select(GEOID, pct_change), by = "GEOID")

# Symmetric diverging breaks centred at 0
brk_max <- 80
mw_map <- mw_map %>%
  mutate(pct_clipped = pmax(pmin(pct_change, brk_max), -brk_max))

ggplot(mw_map) +
  geom_sf(aes(fill = pct_clipped), colour = "white", linewidth = 0.08) +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9641",
    midpoint = 0,
    limits   = c(-brk_max, brk_max),
    breaks   = seq(-brk_max, brk_max, 20),
    labels   = function(x) paste0(ifelse(x > 0, "+", ""), x, "%"),
    name     = "% Change\n2000–2020",
    oob      = scales::squish
  ) +
  coord_sf(datum = NA) +
  labs(
    title    = "County Population Change in the Midwest (2000–2020)",
    subtitle = "IL · IN · IA · KS · MI · MN · MO · NE · ND · OH · SD · WI",
    caption  = "Source: U.S. Decennial Census 2000 & 2020 via tidycensus\nCounties clipped to ±80% for colour scale clarity"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    plot.caption  = element_text(size = 8,  colour = "grey50"),
    legend.position = "right"
  )
```

\newpage

# Summary

## Key Findings – Japan

- **Growing prefectures (2020):** Okinawa and the three major metro areas (Tokyo, Aichi, Osaka) consistently showed positive growth relative to 1995, while rural/northern prefectures (Akita, Aomori, Kochi) declined most sharply.
- **City size:** Larger cities (Metro 700k+) tended to grow or hold steady, while small cities (<30k) disproportionately declined — consistent with urbanisation patterns.
- **Density–growth correlation:** A `r if(exists("r_dens")) round(r_dens,3) else "..."` Pearson correlation between 1995 population density and 1995–2020 % change indicates that denser prefectures generally experienced stronger growth, reflecting urban concentration.

## Key Findings – United States

- **Metro advantage:** Metro counties (RUCC 1–3) showed higher median growth rates than nonmetro counties, but also greater variance driven by Sun Belt metros.
- **Illinois pyramids:** Metro counties show a more working-age-heavy pyramid; nonmetro counties have a relatively older profile with a larger share of 55+ age groups.
- **Midwest choropleth:** Suburban counties surrounding major metros (Chicago, Columbus, Minneapolis) grew substantially, while many rural counties in the Great Plains (North/South Dakota, rural Iowa) experienced decline.

---

*Analysis produced with R `r R.version$major`.`r R.version$minor`; key packages: tidycensus `r packageVersion("tidycensus")`, ggplot2 `r packageVersion("ggplot2")`, sf `r packageVersion("sf")`.*
